---
title: "Reproducible Research: Peer Assessment 1"
auther: Eng.Sarah Hassan
output: 
  html_document:
    keep_md: true
---

```{r echo=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```
Used packages  

```{r}
library(dplyr)
library(xtable)
library(lattice)
```
## Loading and preprocessing the data

```{r}
if(!file.exists("activity.csv")){unzip("activity.zip")}
activityData<-read.csv("activity.csv",header = TRUE)
activityData$date<-as.Date(activityData$date)
```

## What is mean total number of steps taken per day?
```{r}
#totalPerDay<-aggregate(steps~date,data=activityData,FUN = sum,simplify = T,na.rm=T)

summaryPerDay<-activityData %>%
                group_by(date)%>%
                summarise(totalSteps=sum(steps,na.rm=T),
                          meanSteps=mean(steps,na.rm=T),
                          medianSteps=median(steps,na.rm=T),.groups="keep")

plot(summaryPerDay$date,summaryPerDay$totalSteps,xlab="Date",ylab="Total steps",main="Total Steps Per Day",type = "h",lwd=5,lend=1,col="lightblue")
```
```{r results='asis'}

xt<-xtable(summaryPerDay)
print(xt,type="html")

```


## What is the average daily activity pattern?
```{r}
avgStepsPerInterval<-aggregate(steps~interval,data=activityData,FUN = mean,simplify = T,na.rm=T)
plot(avgStepsPerInterval$interval,avgStepsPerInterval$steps,xlab="interval",ylab="Average Steps",main="Average Steps Per Interval ",type = "l",col="lightblue",)

maxStepsInterval<-avgStepsPerInterval[which.max(avgStepsPerInterval$steps),]$interval

```

The 5-minute interval with the maximum steps avereged accross all days is `r maxStepsInterval`

## Imputing missing values

```{r}
countNA<-sum(is.na(activityData$steps))

```

The total number of missing values in the dataset is `r countNA` rows.  
The missing values are replaced by the mean for that 5-minute interval calculated over all days:  
```{r}
updateSteps<-function(df,stepsMeanPerInterval){
        newSteps<-vector(length = nrow(df))
        for(i in seq_len(nrow(df))){
                if(is.na(df[i,"steps"])){
                        newSteps[i]=stepsMeanPerInterval[
                                which(stepsMeanPerInterval$interval==df[i,"interval"]),"steps"]
                }
                else{newSteps[i]=df[i,"steps"]}
        }
        newSteps
}

activityDataNew<-mutate(activityData,steps=updateSteps(activityData,avgStepsPerInterval))
```
Calculaing new totals/mean/median:  
```{r}
summaryPerDayNew<-activityDataNew %>%
                group_by(date)%>%
                summarise(totalSteps=sum(steps,na.rm=T),
                          meanSteps=mean(steps,na.rm=T),
                          medianSteps=median(steps,na.rm=T),.groups="keep")

plot(summaryPerDayNew$date,summaryPerDayNew$totalSteps,xlab="Date",ylab="Total steps",main="Total Steps Per Day",type = "h",lwd=5,lend=1,col="lightblue")
```
```{r results='asis'}

xtNew<-xtable(summaryPerDayNew)
print(xtNew,type="html")

```

Comparing the two plots (before and after removing NAs):  
```{r}
par(mfcol=c(1,2))
plot(summaryPerDay$date,summaryPerDay$totalSteps,xlab="Date",ylab="Total steps",main="Total Steps Per Day",type = "h",lwd=5,lend=1,col="lightblue")
plot(summaryPerDayNew$date,summaryPerDayNew$totalSteps,xlab="Date",ylab="Total steps",main="Total Steps Per Day",type = "h",lwd=5,lend=1,col="lightblue")
```
The impact of imputing missing data on the estimates of the total daily number of steps: it helped filling the gaps by eliminating NAs without afftecting summaries  

## Are there differences in activity patterns between weekdays and weekends?

```{r}
activityDataNew<-mutate(activityDataNew,dayType=
                     factor( weekdays(activityDataNew$date,T)%in% c("Sat","Sun"),
                             labels =c("weekday","weekend"))
                     )
```

```{r}
avgStepsPerIntervalNew<-aggregate(steps~interval+dayType,data=activityDataNew,FUN = mean,simplify = T,na.rm=T)


xyplot(steps~interval|dayType,data = avgStepsPerIntervalNew,type="l",layout=c(1,2))
```

